---
title: "On Statistical Power"
output: word_document
---

Statistical Power in Binary Logistic Regression

$log(odds) = log\frac{p}{1-p} = \beta_0 + \beta_1x$

$H_0: \beta_1=0$
$H_1: \beta_1\neq0$

Determining whether a statistical test has the power to detect an odds ratio for an outcome of interest (eg: Cancer incidence).

$n = \frac{(Z_{1-\alpha/2}\sqrt{V_0}+Z_P\sqrt{V})^2}{(\beta-\beta_0)^2}$
$n = \frac{(Z_{1-\alpha/2}\sqrt{V_0}+Z_P\sqrt{V})^2}{P1(1-P1)\beta^{*2}}$
where $\alpha$ = significance level, and $P$ = power

$H_0: \beta=\beta_0$
$H_1: \beta\neq\beta_0$i

```{r}
gwas_cancer_data %>% dplyr::select(RISK.ALLELE.FREQUENCY) %>% head
    colnames
```


```{r}
breast_pubs %>% 
    mutate(
        init_sample = as.numeric(init_cases) + as.numeric(init_controls),
        rep_sample  = as.numeric(rep_cases)  + as.numeric(rep_controls)
    ) %>% 
    dplyr::select(
        PUBMEDID,
        init_cases,
        init_controls,
        init_sample,
        P.VALUE,
        OR.or.BETA,
        lower_ci,
        upper_ci,
        rep_cases,
        rep_controls,
        rep_sample
    ) %>% 
    head()
```

```{r}
wp.logistic(754, 0.1, 0.25, 0.05, alternative = "greater",family = "Bernoulli")
```
```{r}
powerLogisticBin(754, 0.1, 0.25, alpha = 0.05, B)
```

